---
title: "Trends in Opioid Prescriptions"
author: "Kyle Brewster"
date: '2022-04-06'
output:
  ioslides_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## The Big Picture

## The Data

* Source
    - [Kaggle](https://www.kaggle.com/datasets/apryor6/us-opiate-prescriptions?select=prescriber-info.csv) data set sourced from [CMS.gov](https://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/Medicare-Provider-Charge-Data/Part-D-Prescriber) (Centers for Medicare & Medicaid Services)
* Challenges
    - Thousands of spelling inconsistencies 
    - Sifting through extra/unnecessary data included with download
    - Difficulty grouping states by type of statewide-opioid policies
    - Potential for confounding variables outside of data set
    
## The Data (continued)
* Shortcomings
    - No obvious predictor/outcome variables
    - Only contains data collected in 2014
    - Large number of observations requires higher computational demand
    - Lots of interaction among different variables, could skew model if
      improperly handled
    

## The Data Visualized
```{r map_of_deaths}
map_of_deaths
```

## Surprises from the Data

Since the observations in the data were majority male providers, it was not a
surprise to see that most of the prescriptions written were done so by male
providers.
```{r}

```

It *was* surprising to see the difference between prescriptions rates once
controlling for the difference in gender
```{r}

```


## Results & Conclusions

One of the models was able to obtain an `MSE` approximately equal to zero, but
was because of interdependence of variables selected for model (i.e. the model
would perform poorly if number of opioids prescribed was withheld from data)

The best performing model was *__Model #__*

Taeawakes?
- 

```{r pressure}
plot(pressure)
```

